---
title: "cvd_gender"
author: "Sarah Zhang"
date: '2022-09-07'
output: html_document
---

Notes: crude rate = death count / population * 100000

Import data: monthly CVD deaths by sex in 2018, 2019, and 2020 
```{r}
cvd_age = read.csv("cvd_age_cleaned.csv")
```

Regroup age groups 
```{r}
ageGroup = lm(crude_rate ~ age_group , data = cvd_age)
summary(ageGroup)
```
**Answer:** Based on p-values, regrouping the age group into three groups: 05-44, 45-64, 65+. Data is re-cleaned in python to file cvd_age_regrouped


```{r}
cvdAgeRegroup = read.csv('cvd_age_regrouped.csv')
```

a. Add time indicator by month
```{r}
cvdAgeRegroup = transform(cvdAgeRegroup, month_code = substr(month,1,2))
```

```{r}
age_month_model = lm(crude_rate ~ age_group*month_code, data = cvdAgeRegroup)
summary(age_month_model)
```

**Analysis: **

b. Add time indicator on COVID:  
        - '0 represents period before COVID
        - '1' represents period after COVID
```{r}
cvdAgeRegroup$covid = ifelse(grepl('20', cvdAgeRegroup$month)== TRUE, 1, 0)
```

c. Generate modeling results
```{r}
age_covid_model = lm(crude_rate ~ age_group*covid, data = cvdAgeRegroup)
summary(age_covid_model)
```
**Analysis:**


b. Add time indicator on seasonality.
According to 50 line plot of crude rate by states and sex_month_model result, CVD death rate reach to a **high level** every year between **November and April next year**; and CVD death rate reach to a **low point** every year between **May to October**. In order to further test significance of seasonality on CVD crude rate, we add a time indicator of 'season' based on months with high and low crude rate. 
  -- 'high' represents months with high crude rate 
  -- 'low' represents months with low crude rate 
```{r}
cvdAgeRegroup$month_code = as.numeric(cvdAgeRegroup$month_code)
```
  
```{r}
cvdAgeRegroup = transform(cvdAgeRegroup, 
                          season= ifelse(month_code %in% c(5:10),'low', 'high'))
```
  
```{r}
cvdAgeRegroup$month_code = toString(cvdAgeRegroup$month_code)
```

