---
title: "CVD_sex"
author: "Sarah Zhang"
date: '2022-07-11'
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r}
library("ggplot2")
library(MASS)
```

## R Markdown

Notes: crude rate = death count / population * 100000

1. Testing significance of factors 'sex' and 'time_indicators' to CVD crude rate. 

a. Import data: monthly CVD deaths by sex in 2018, 2019, and 2020 
```{r}
cvd_sex = read.csv("cvd_sex_cleaned.csv")
```

a. Add time indicator by month
```{r}
month = rep(rep(1:12, each=2),3)
cvd_sex$month_code = as.character(month)
```

b. Generate modeling result
```{r}
sex_month_model = lm(crude_rate ~ gender*month_code, data = cvd_sex)
summary(sex_month_model)
```
**Analysis: **



b. Add time indicator on COVID:  
        - '0 represents period before COVID
        - '1' represents period after COVID
```{r}
cvd_sex$covid = ifelse(grepl('20', cvd_sex$month)== TRUE, 1, 0)
```

c. Generate modeling results
```{r}
sex_covid_model = lm(crude_rate ~ gender*covid, data = cvd_sex)
summary(sex_covid_model)
```

**Analysis: **



b. Add time indicator on seasonality.
According to 50 line plot of crude rate by states and sex_month_model result, CVD death rate reach to a high level every year between November and April next year; and CVD death rate reach to a low point every year between May to October. In order to further test significance of seasonality on CVD crude rate, we add a time indicator of 'season' based on months with high and low crude rate. 
  -- 'high' represents months with high crude rate 
  -- 'low' represents months with low crude rate 
```{r}
high18 = rep('high',(match('04/1/18',cvd_sex$month)-match('01/1/18',cvd_sex$month)+2))
low18 = rep('low',(match('10/1/18',cvd_sex$month)-match('05/1/18',cvd_sex$month)+2))
high19 = rep('high',match('04/1/19',cvd_sex$month)-match('11/1/18',cvd_sex$month)+2)
low19 = rep('low',match('10/1/19',cvd_sex$month)-match('05/1/19',cvd_sex$month)+2)
high20 = rep('high',match('04/1/20',cvd_sex$month)-match('11/1/19',cvd_sex$month)+2)
low20 = rep('low',match('10/1/20',cvd_sex$month)-match('05/1/20',cvd_sex$month)+2)
high21 = rep('high',match('12/1/20',cvd_sex$month)-match('11/1/20',cvd_sex$month)+2)
season = append(append(append(append(append(append(high18,low18),high19),low19),high20),low20),high21)
cvd_sex$season = season
```

c. Generate Modeling result
```{r}
sex_season_model = lm(crude_rate ~ gender*season, data = cvd_sex)
summary(sex_season_model)
```

**Analysis:**


b. Add a time indicator by year (2018, 2019, 2020)
```{r}
y2018 = rep('2018',match('01/1/19',cvd_sex$month)-match('01/1/18',cvd_sex$month))
y2019 = rep('2019',(match('01/1/20',cvd_sex$month)-match('01/1/19',cvd_sex$month)))
y2020 = rep('2020',(match('12/1/20',cvd_sex$month)-match('01/1/20',cvd_sex$month)+2))
year = append(append(y2018,y2019),y2020)
cvd_sex$year = year
```

c. Generate modeling result
```{r}
sex_year_model = lm(crude_rate ~ gender*year, data = cvd_sex)
summary(sex_year_model)
sex_year_model
```

d. Build a full model with gender and all time indicators 
```{r}
sex_time_model = lm(crude_rate ~ gender + covid + year + season + month_code , data = cvd_sex)
summary(sex_time_model)
```
```{r}
sex_time_step.both = step(sex_time_model, direction='both')
summary(sex_time_step.both)
```

```{r}
cvd_age = read.csv('cvd_age_cleaned.csv')
```

```{r}
age_model = lm(crude_rate ~ age_group, data = cvd_age)
summary(age_model)
```
```{r}
cvd_age = read.csv('cvd_age_cleaned.csv')
```


