---
title: "cvd_region_confounder"
author: "Zhongyi Guo"
output: pdf_document
---

```{r}
suppressWarnings(library(tidyverse))
suppressWarnings(library(ggplot2))
suppressWarnings(library(rstatix))
```

## Exploratory Data Analysis

```{r}
# load region data
region <- read.csv("../data/cvd_region_crude_rate.csv")

# create a boolean dummy column named `before_covid` to indicate if this row is 
# before COVID. If yes, before_covid = 1; if no, before_covid = 0
before_covid_index <- append(grep("2018", region$month), grep("2019", region$month))
region$before_covid <- 0
region$before_covid[before_covid_index] <- 1

# report summary statistics of `crude_rate`
region %>%
  group_by(region) %>%
  summarise(
    count = n(),
    mean = mean(crude_rate, na.rm = TRUE),
    sd = sd(crude_rate, na.rm = TRUE),
    median = median(crude_rate, na.rm = TRUE),
    IQR = IQR(crude_rate, na.rm = TRUE)
    )
```

```{r}
ggplot(region, aes(x = region, y = crude_rate)) + geom_boxplot() +
  labs(x = "region", y = "crude rate", title = "Boxplot of crude rate by region") +
  theme_minimal()
```

The boxplot indicates some outliers in all regions. We will save them for now 
and remove them later.

```{r}
region %>%
  group_by(region) %>%
  ggplot(aes(x=crude_rate)) + 
  geom_histogram() + facet_wrap(~ region, nrow = 2, ncol =2)
```

## Normality test

```{r}
shapiro.test(region$crude_rate)
```

The p-value of the Shapiro normality test is 8.787 x 10^(-12), which indicates 
that the crude rate of all regions is not normally distributed. Thus, for the 
next step, we will detect and remove outliers of each region.

## Outlier Removal

First, we want to observe the mean of the crude rate in each region.

```{r}
region %>%
 group_by(region) %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

Second, we want to identify outliers and remove.

```{r}
region_outliers <- region %>%
  group_by(region) %>%
  identify_outliers(crude_rate)
region_outliers
```

```{r}
region <- region %>%
  anti_join(region_outliers)
```

```{r}
region %>%
  group_by(region) %>%
  shapiro_test(crude_rate)
```

The result shows that the crude rate of each region is not normally distributed.

```{r}
region %>%
  levene_test(crude_rate ~ region)
```

The p-value of the Levene test is 0.4936763. It indicates that there is no enough 
evidence to reject the null hypothesis that each region has equal variance.

## ANOVA - Are the means of crude rate in each region different?

```{r}
anova_result <- region %>%
  aov(crude_rate ~ region, data = .)
summary(anova_result)
```

The ANOVA test of region returns a p-value that is smaller than <2 x 10^(-16), 
very close to 0. We are confident to reject the null hypothesis of the ANOVA test.
Thus, the differences in means of crude rate between the groups are statistically significant.

```{r}
ggplot(data = region)+
  geom_boxplot( mapping = aes(x = region, y = crude_rate)) +
  labs(x = "Pre/Post COVID", y = "Crude rates", title = "Crude Rates Pre- and Post- COVID")+ 
  theme(plot.title = element_text(hjust = 0.5))
```


## Pre- vs. Post-COVID

In this section, we want to separate data into two groups: Pre- and Post-COVID. 
We want to study if there is statistical significant evidence that indicates the
means of crude rate in each region are different.

### Northeast

```{r}
northeast <- region[region$region %in% "Northeast", ]
head(northeast)
```

```{r}
northeast %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
northeast %>%
  t_test(crude_rate ~ before_covid, alternative = "less") %>%
  add_significance()
```
p-value of this test is 0.941, which fails to reject the null hypotheis. The 
means of crude rate are different in the Northeastern region Pre- vs. Post-COVID.

```{r}
boxplot(crude_rate ~ before_covid, data = northeast, xlab = "Pre/Post COVID", 
        ylab = "Crude Rate", main = "Crude Rates Pre- and Post- COVID in the Northeastern region")
```

### Midwest

```{r}
midwest <- region[region$region %in% "Midwest", ]
head(midwest)
```

```{r}
midwest %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
midwest %>%
  t_test(crude_rate ~ before_covid, alternative = "less") %>%
  add_significance()
```

p-value of this test is 1, which fails to reject the null hypotheis. The 
means of crude rate are different in the Midwestern region Pre- vs. Post-COVID.

```{r}
boxplot(crude_rate ~ before_covid, data = midwest, xlab = "Pre/Post COVID", 
        ylab = "Crude Rate", main = "Crude Rates Pre- and Post- COVID in the Midwestern region")
```

### South

```{r}
south <- region[region$region %in% "South", ]
head(south)
```

```{r}
south %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
south %>%
  t_test(crude_rate ~ before_covid, alternative = "less") %>%
  add_significance()
```

p-value of this test is 1, which fails to reject the null hypotheis. The 
means of crude rate are different in the Southern region Pre- vs. Post-COVID.

```{r}
boxplot(crude_rate ~ before_covid, data = south, xlab = "Pre/Post COVID", 
        ylab = "Crude Rate", main = "Crude Rates Pre- and Post- COVID in the Southern region")
```

### West

```{r}
west <- region[region$region %in% "West", ]
head(west)
```

```{r}
west %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
west %>%
  t_test(crude_rate ~ before_covid, alternative = "less") %>%
  add_significance()
```

p-value of this test is 1, which fails to reject the null hypotheis. The 
means of crude rate are different in the Southern region Pre- vs. Post-COVID.

```{r}
boxplot(crude_rate ~ before_covid, data = west, xlab = "Pre/Post COVID", 
        ylab = "Crude Rate", main = "Crude Rates Pre- and Post- COVID in the Western region")
```
