---
title: "cvd_region_confounder"
author: "Zhongyi Guo"
output: pdf_document
---

```{r}
#| label: load_package
library(tidyverse)
library(ggplot2)
library(rstatix)
```

## Exploratory Data Analysis

```{r}
#| label: load_clean_data
# load region data
region <- read.csv("../data/cvd_region_crude_rate.csv")

# create a dummy variable `covid` to indicate if this row of data is 
# before COVID or not. If before COVID happened (Pre-COVID), covid = 0; if after 
# COVID happened, covid = 1
before_covid_index <- append(grep("2018", region$month), grep("2019", 
                                                              region$month))
region$covid <- 1
region$covid[before_covid_index] <- 0
```

```{r}
#| label: summary_statistics
# report summary statistics of `crude_rate`
region %>%
  group_by(region) %>%
  summarise(
    count = n(),
    mean = mean(crude_rate, na.rm = TRUE),
    sd = sd(crude_rate, na.rm = TRUE),
    median = median(crude_rate, na.rm = TRUE),
    IQR = IQR(crude_rate, na.rm = TRUE)
    )
```

```{r}
#| label: boxplot_of_crude_rate_for_each_region
# create a boxplot for crude rates of each region 
ggplot(region, aes(x = region, y = crude_rate)) + geom_boxplot() +
  labs(x = "region", y = "crude rate", 
       title = "Boxplot of crude rate by region") + theme_bw()
```

The boxplot indicates some outliers in all regions. We will save them for now 
and remove them later.

```{r}
#| label: histogram_of_crude_rate_for_each_region
# create histograms for each region, wrapped in facet
region %>%
  group_by(region) %>%
  ggplot(aes(x=crude_rate)) + geom_histogram(binwidth = 10) + 
  facet_wrap(~ region, nrow = 2, ncol = 2) +
  labs(x = "crude rate") +
  theme_bw()
```

### Normality test

```{r}
#| label: normality_test_of_crude_rate_for_each_region
# report p-values of Shapiro test for each region
region %>%
  group_by(region) %>%
  summarize(p_value = shapiro.test(crude_rate)$p.value)
```

The p-value of the Shapiro normality test for each region are all smaller than 
0.05, which indicates that the crude rate of all regions is not normally 
distributed. Thus, for the next step, we will first detect and remove 
outliers of each region.

### Outlier Removal

```{r}
#| label: find_outliers_of_each_region
# find outliers of each region
region_outliers <- region %>%
  group_by(region) %>%
  identify_outliers(crude_rate)
region_outliers
```

```{r}
#| label: anti_join_to_remove_ourliers
# remove outliers for each region
region <- region %>%
  anti_join(region_outliers)
```

```{r}
#| label: histogram_of_crude_rate_for_each_region_without_outliers
# create histograms for each region, wrapped in facet
region %>%
  group_by(region) %>%
  ggplot(aes(x=crude_rate)) + geom_histogram(binwidth = 5) + 
  facet_wrap(~ region, nrow = 2, ncol = 2) + 
  labs(x = "crude rate") +
  theme_bw()
```

```{r}
#| label: normality_test_of_crude_rate_for_each_region_without_outliers
# perform normality test again after outlier removal
region %>%
  group_by(region) %>%
  shapiro_test(crude_rate)
```

The result shows that the crude rate of each region is not normally distributed, 
after outlier removal. We will then perform a Wilcox test for rank sum.

```{r}
#| label: wilcox_test_for_distribution_comparison
pairwise.wilcox.test(region$crude_rate, region$region, p.adjust.method = "none")
```

We can conclude Midwest and Northeast, Midwest and South, Midwest and West, 
Northeast and West, and South and West have strong evidence to suggest that 
these groups have significantly different distributions because of smaller than
0.05 p-values. There is insufficient evidence to conclude that there is a 
significant difference between Northeast and South.

This observation can be further explored in discussion with regards to healthcare 
systems, weather, economic levels, etc., of each region.

Now, the dataset `region` is clean. We will save it as a new file named 
`cvd_region_crude_rate.csv` and reimport it.

```{r}
write.csv(region, file = "../data/cvd_region_crude_rate_clean.csv", 
          row.names = FALSE)
region <- read.csv("../data/cvd_region_crude_rate_clean.csv")
```

## ANOVA - Are crude rate means different among regions?

```{r}
#| label: anova_crude_rate
anova_result <- region %>%
  aov(crude_rate ~ region, data = .)
summary(anova_result)
```

The ANOVA test of region returns a p-value that is smaller than 2 x 10^(-16), 
very close to 0. We are confident to reject the null hypothesis of the ANOVA test.
Thus, the differences in means of crude rate between each region are 
statistically significant.

## Pre- vs. Post-COVID Analysis

In this section, we want to separate data into two groups: Pre- and Post-COVID. 
We want to study if there is statistical significant evidence that indicates the
means of crude rate in each region are different.

### General

```{r}
#| label: pre_post_COVID_visualizatoin
region %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  ggplot(aes(x = region, y = crude_rate)) + geom_boxplot() +
  facet_wrap(~ factor(covid, levels = c("Pre-COVID", "Post-COVID")), nrow = 1) +
  labs(x = "region", y = "crude rate") +
  ggtitle("Pre- and Post-COVID boxplot of crude rate") +
  theme_bw()
```

```{r}
#| label: pre_post_COVID_mean_comparison
pre_post_crude_rate_mean <- region %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>%
  summarize(mean_value = mean(crude_rate))
pre_post_crude_rate_mean
```

```{r}
region %>% 
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>% ggplot(aes(x=crude_rate)) + geom_histogram(binwidth = 5) +
  facet_wrap(~factor(covid, levels = c("Pre-COVID", "Post-COVID")), nrow = 2) +
  theme_bw() 
```

The crude rate means Pre- and Post-COVID are different in general.

### Each region

#### Northeast

```{r}
#| label: northeastern_data_extraction
northeast <- region[region$region %in% "Northeast", ]
```

```{r}
#| label: northeastern_summary_statistics
northeast %>%
  get_summary_stats(crude_rate, type = "mean_sd")
```

```{r}
#| label: northeastern_pre_post_COVID_crude_rate_variance
northeast %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>%
  summarise(var = var(crude_rate))
```

Because the variances of crude rate Pre- and Post-COVID are not close, we will 
set the parameter `var.equal` = FALSE.

```{r}
#| label: northeastern_pre_post_COVID_t_test
northeast %>%
  t_test(crude_rate ~ covid, alternative = "less", var.equal = FALSE) %>%
  add_significance()
```

p-value of this test is 0.0591, which fails to reject the null hypothesis. The 
crude rate means in the Northeastern region are statistically same between 
Pre- and Post-COVID.

```{r}
#| label: northeastern_pre_post_COVID_boxplot
northeast %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  ggplot(aes(x = factor(covid, levels = c("Pre-COVID", "Post-COVID")), 
             y = crude_rate)) + geom_boxplot() +
  labs(x = "time", y = "crude rate") +
  ggtitle("Pre- and Post-COVID crude rate boxplot in the Northeastern region") +
  theme_bw()
```

From the boxplot we observed that the means of crude rate Pre-COVID and Post-COVID 
are very close in the Northeastern region, which matched the conclusion from the 
t test.

#### Midwest

```{r}
#| label: midwestern_data_extraction
midwest <- region[region$region %in% "Midwest", ]
```

```{r}
#| label: midwestern_summary_statistics
midwest %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
#| label: northeastern_pre_post_COVID_crude_rate_variance
midwest %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>%
  summarise(var = var(crude_rate))
```

Because the variances of crude rate Pre- and Post-COVID are not close, we will 
set the parameter `var.equal` = FALSE.


```{r}
#| label: midwestern_pre_post_COVID_t_test
midwest %>%
  t_test(crude_rate ~ covid, alternative = "less", var.equal = FALSE) %>%
  add_significance()
```

The p-value of the Midwestern Pre- and Post-COVID t test on crude rate means is 
0.000359, which indicates that the null hypothesis should be rejected. The crude 
rate means are statistically different in the Midwestern region Pre- vs. 
Post-COVID.

```{r}
#| label: midwestern_pre_post_COVID_boxplot
midwest %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  ggplot(aes(x = factor(covid, levels = c("Pre-COVID", "Post-COVID")), 
             y = crude_rate)) +
  geom_boxplot() + labs(x = "time", y = "crude rate") +
  ggtitle("Pre- and Post-COVID crude rate boxplot in the Midwestern region") +
  theme_bw()
```

From the boxplot, we observed the crude rate means are different Pre- vs. Post-
COVID in the Midwestern region, which matched the conclusion from the t test.

#### South

```{r}
#| label: southern_data_extraction
south <- region[region$region %in% "South", ]
```

```{r}
#| label: southern_summary_statistics
south %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
#| label: southern_pre_post_COVID_crude_rate_variance
south %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>%
  summarise(var = var(crude_rate))
```

Because the variances of crude rate Pre- and Post-COVID are not close, we will 
set the parameter `var.equal` = FALSE.

```{r}
#| label: southern_pre_post_COVID_t_test
south %>%
  t_test(crude_rate ~ covid, alternative = "less", var.equal = FALSE) %>%
  add_significance()
```

The p-value of the Southern Pre- and Post-COVID t test on crude rate means is 
5.37 x 10^(-5), which indicates that the null hypothesis should be rejected. The
crude rate means are statistically different in the Southern region Pre- vs. 
Post-COVID.

```{r}
#| label: southern_pre_post_COVID_boxplot
south %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  ggplot(aes(x = factor(covid, levels = c("Pre-COVID", "Post-COVID")), 
             y = crude_rate)) +
  geom_boxplot() + labs(x = "time", y = "crude rate") +
  ggtitle("Pre- and Post-COVID crude rate boxplot in the Southern region") +
  theme_bw()
```

From the boxplot, we observed the crude rate means are different Pre- vs. Post-
COVID in the Southern region, which matched the conclusion from the t test.

#### West

```{r}
#| label: western_data_extraction
west <- region[region$region %in% "West", ]
```

```{r}
#| label: western_summary_statistics
west %>%
  get_summary_stats(crude_rate,type = "mean_sd")
```

```{r}
#| label: western_pre_post_COVID_crude_rate_variance
west %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  group_by(covid) %>%
  summarise(var = var(crude_rate))
```

The variances are close. We will set `var.equal` = TRUE in the t test.

```{r}
#| label: western_pre_post_COVID_t_test
west %>%
  t_test(crude_rate ~ covid, alternative = "less", var.equal = TRUE) %>%
  add_significance()
```

The p-value of the Western Pre- and Post-COVID t test on crude rate means is 
4.97 x 10^(-6), which indicates that the null hypothesis should be rejected. The
crude rate means are statistically different in the Western region Pre- vs. 
Post-COVID.

```{r}
#| label: western_pre_post_COVID_boxplot
west %>%
  mutate(covid = recode(covid, "0" = "Pre-COVID", "1" = "Post-COVID")) %>%
  ggplot(aes(x = factor(covid, levels = c("Pre-COVID", "Post-COVID")), 
             y = crude_rate)) +
  geom_boxplot() + labs(x = "time", y = "crude rate") +
  ggtitle("Pre- and Post-COVID crude rate boxplot in the Western region") +
  theme_bw()
```

From the boxplot, we observed the crude rate means are different Pre- vs. Post-
COVID in the Southern region, which matched the conclusion from the t test.

## Conclusion

1. The ANOVA result concludes that crude rate means are different among all 
regions.
2. The crude rate means Pre- and Post-COVID are different among all regions 
in general.
3. Midwestern, Soutern, Western regions display statistically significant 
difference in crude rate means Pre- and Post-COVID, while the Northeastern region 
does not.
